---
title: "water_project"
author: "Armen Mkrtumyan"
date: "2024-11-20"
output: pdf_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

```{r}
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(imputeTS)
library(dplyr)
library(readxl)
library(sf)
library(ggplot2)
library(shiny)
library(lubridate)
library(plotly)
library(reshape2)
library(terra)
library(tmap)
library(mapview)
```

```{r}
data <- read_excel("./Data/water_data2.xlsx")
```

```{r}
boundary_coords <- data.frame(
  Name = c("upper left 1", "upper left 2", "upper left middle",
           "lower left 1", "lower left 2",
           "upper right 1", "upper right 2", "upper right middle",
           "lower right 1", "lower right 2"),
  Y = c(40.596992, 40.552921, 40.517957,
        40.427392, 40.434448,
        40.626964, 40.589431, 40.537007,
        40.500207, 40.459469),
  X = c(44.967396, 44.999669, 44.953664,
        45.109524, 45.209088,
        44.997609, 45.103009, 45.206349,
        45.245145, 45.264714)
)

print(boundary_coords)

```

Lets fill out those X and Y values by using Google Map coordinates

```{r}
centroid_X <- mean(boundary_coords$X)
centroid_Y <- mean(boundary_coords$Y)

boundary_coords <- boundary_coords %>%
  mutate(
    Angle = atan2(Y - centroid_Y, X - centroid_X) * (180 / pi)
  )

boundary_coords_ordered <- boundary_coords %>%
  arrange(Angle)

boundary_coords_ordered <- boundary_coords_ordered %>%
  add_row(Name = boundary_coords_ordered$Name[1],
          Y = boundary_coords_ordered$Y[1],
          X = boundary_coords_ordered$X[1],
          Angle = boundary_coords_ordered$Angle[1])

print(boundary_coords_ordered)

```

```{r}
boundary_matrix <- as.matrix(boundary_coords_ordered[, c("X", "Y")])
boundary_polygon <- st_polygon(list(boundary_matrix))
boundary_sf <- st_sfc(boundary_polygon, crs = 4326)
st_is_valid(boundary_sf)

```

```{r}
small_sevan_indices <- which(data$Water_Object == "Սևանա լիճ - Փոքր Սևան" | grepl("Փոքր Սևան", data$Station_Location))
num_small_sevan <- length(small_sevan_indices)

cat("Number of 'Small Sevan' rows to impute:", num_small_sevan, "\n")
if (num_small_sevan != 261) {
  warning("Expected 261 'Small Sevan' rows, but found ", num_small_sevan)
}
```


```{r}
set.seed(123)
sampled_points <- st_sample(boundary_sf, size = num_small_sevan, type = "random")

if (length(sampled_points) < num_small_sevan) {
  stop("Not enough points could be sampled within the polygon.")
}

sampled_coords <- st_coordinates(sampled_points)
sampled_coords_df <- as.data.frame(sampled_coords)
colnames(sampled_coords_df) <- c("X", "Y")
head(sampled_coords_df)

```

```{r}
data_imputed <- data

data_imputed$X[small_sevan_indices] <- sampled_coords_df$X
data_imputed$Y[small_sevan_indices] <- sampled_coords_df$Y

na_X <- sum(is.na(data_imputed$X[small_sevan_indices]))
na_Y <- sum(is.na(data_imputed$Y[small_sevan_indices]))

cat("Number of NA X for 'Small Sevan':", na_X, "\n")
cat("Number of NA Y for 'Small Sevan':", na_Y, "\n")

if (na_X > 0 | na_Y > 0) {
  warning("There are still NA values in 'Small Sevan' coordinates.")
}

```

```{r}
numeric_cols <- data_imputed %>%
  select(
    Dissolved_Oxygen, BOD5, COD,
    NH4, NO2, NO3, PO4,
    Zn, Cu, Cr, As, Cd, Pb, Ni, Mo, Mn, V, Co, Fe,
    Ca, Mg, Ba, Be, K, Na,
    B, Al, Se, Sb, Sn,
    Total_Inorganic_N_mg_N_L,
    P, Cl, SO4, Si,
    TDS, EC_microS_cm,
    Hardness, Suspended_Solids,
    X, Y
  ) %>%
  names()

length(numeric_cols)
numeric_data <- data_imputed %>%
  select(all_of(numeric_cols))

```


# Handling missing data

Linear interpolation is a good choice for numeric data where the assumption of a smooth progression between points makes sense (e.g., time-series data or other continuous data).

```{r}
na_counts <- colSums(is.na(numeric_data))
na_counts_df <- data.frame(Column = names(na_counts), NA_Count = na_counts)

na_counts_df <- na_counts_df[order(-na_counts_df$NA_Count), ]
print(na_counts_df)
```

Including X and Y in PCA:
Pros: Incorporates spatial information into the PCA, potentially revealing spatial patterns.
Cons: Spatial coordinates can dominate the variance, possibly overshadowing water quality parameters.

```{r}
numeric_scaled <- scale(numeric_data)

pca_result <- prcomp(numeric_scaled, center = TRUE, scale. = TRUE)
summary(pca_result)

fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Variance Explained by Principal Components")

pca_scores <- as.data.frame(pca_result$x)
data_pca <- data_imputed %>%
  bind_cols(pca_scores)

fviz_nbclust(numeric_scaled, kmeans, method = "wss") +
  ggtitle("Elbow Method for Determining Optimal Clusters") +
  scale_y_continuous(labels = scales::comma_format()) +
  theme_minimal()

optimal_k <- 3

set.seed(123)
kmeans_result <- kmeans(numeric_scaled, centers = optimal_k, nstart = 25)

data_pca$Cluster <- as.factor(kmeans_result$cluster)

fviz_pca_ind(pca_result,
             geom.ind = "point",
             habillage = data_pca$Cluster,
             palette = "jco",
             addEllipses = TRUE,
             legend.title = "Cluster") +
  ggtitle(paste("PCA Clustering (k =", optimal_k, ")"))
```

```{r}
cluster_region_table <- table(data_pca$Cluster, data_pca$Region)
print(cluster_region_table)

fviz_pca_biplot(
  pca_result, 
  geom.ind = "point", 
  pointshape = 21, 
  pointsize = 2,
  fill.ind = data_pca$Cluster,
  col.var = "black",
  arrowsize = 0.8,
  repel = TRUE
) +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73")) +
  labs(title = "PCA Biplot with Clusters", fill = "Cluster") +
  theme_minimal()


ggsave("PCA_Biplot.pdf", width = 8, height = 6)

```

Cluster 1: Higher pollution indicators (e.g., NH4, EC).
Cluster 3: Better water quality (high DO).

Dissolved_Oxygen, Pb, and Zn play a significant role in defining the variability in water quality parameters.
NH4 and EC_microS_cm are important but in the opposite direction, potentially related to higher pollution or industrial impact.

ADD GEOM GITTER3

```{r}

elevation_data <- rast("./Data/ArmDEM.tif")

elevation_data <- setMinMax(elevation_data)

elevation_data_agg <- aggregate(elevation_data, fact = 2, fun = mean)

cat("Original resolution:", res(elevation_data), "\n")
cat("Aggregated resolution:", res(elevation_data_agg), "\n")


```

```{r}
elevation_data <- rast("./Data/ArmDEM.tif")
elevation_data <- setMinMax(elevation_data)

aggregation_factor <- 16
elevation_data_agg <- aggregate(elevation_data, fact = aggregation_factor, fun = mean)

print(object.size(elevation_data_agg), units = "Mb")

interactive_map <- mapview(
  elevation_data_agg, 
  legend = FALSE, 
  layer.name = "Elevation",
  col.regions = terrain.colors(10)
)

interactive_map

if (!exists("interactive_map")) {
  tmap_mode("view")
  tm <- tm_shape(elevation_data_agg) +
    tm_raster(
      palette = terrain.colors(10),
      title = "Elevation",
      popup.vars = "Elevation"
    ) +
    tm_layout(legend.outside = TRUE)
  
  tm
}

tmap_mode("plot")
tm_static <- tm_shape(elevation_data_agg) +
  tm_raster(
    palette = terrain.colors(10),
    title = "Elevation"
  ) +
  tm_layout(legend.outside = TRUE)

```

